Synthesis of Linear Functional Programs

### 2021-07-15 article/article.tex
Introduction first iteration

% \usemintedstyle{lovelace}
\usemintedstyle{murphy}

\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita - Nº 55902 \\
    Supervisor: Bernardo Toninho}
\date{July 2021}
\institute{NOVA School of Science and Technology}

@@

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program (i.e.~generating code) from a high-level specification.
%
Synthesis can be seen as a means to improve a programmer's productivity and/or
program correctness (e.g. through suggestion and/or autocompletion), or as a
tool to automate certain parts of the programming process, to name a few
examples.
% TODO: Escrever um início melhor

Specifications can take many forms (e.g.  polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind of
specification, program synthesizers have to deal with two main inherent sources
of complexity: searching over a vast space of (potentially) valid programs, and
interpreting user intent.
%
In this work we'll explore type-based synthesis where specifications take the
form of linear types, which limit resource usage in programs, in combination
with polymorphism and refinements.  Concretely, we will explore the proof-theoretic foundations
of linear types via the Curry-Howard correspondence with linear logic, viewing
program synthesis under the lens of (linear logic) bottom-up proof search.

%

\mypara{Type-Based Synthesis}
% Type based synthesis
Type-based synthesis uses types as a form of program specification,
automatically or semi-automatically producing a program expression
with the given type and so matching the specification.
%
In order to make type-based specifications more expressive, type-based
synthesis frameworks use rich type systems such as refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16} and polymorphic types~\cite{DBLP:conf/pldi/PolikarpovaKS16}.

Richer type systems allow for more precise types, which can
statically eliminate various kinds of logical errors by making certain
invalid program states ill-typed (e.g.,~a ``null aware'' type system,
will ensure at compile-time that you cannot dereference a
null-pointer). However, they can also be a burden -- the whole point
of these type systems is to ensure that \emph{fewer} programs are
deemed well-typed, which can pose additional challenges to the
development process. Type-based synthesis leverages rich types as a
way of pruning the search space, and by using types the user 
is given a more ``familiar'' specification interface. For instance, the type
$\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
specifies a (curried) function that takes two integers and produces an
integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).
However, the richer type $(x{:}\mathsf{Int}) \rightarrow
(y{:}\mathsf{Int}) \rightarrow \{z{:}\mathsf{Int} \mid z = x+y\}$
very precisely specifies a function that takes integer arguments $x$
and $y$ and returns an integer $z$ that is the sum of $x$ and $y$.

%
This work explores the synthesis of linear functional
programs from types based on linear logic propositions (i.e.~linear
types) by leveraging the Curry-Howard correspondence.

@@

Linear types differ from more traditional types in
that they constrain resource usage in programs by \emph{statically}
limiting the number of times certain resources can be used during
their lifetime.  They can be applied to resource-aware programming
such as concurrent programming (e.g. session types for message passing
concurrency~\cite{DBLP:journals/mscs/CairesPT16}), and to memory-management (e.g.~Rust's ownership
types).

@@

\section{Overview}

Escrever introdução melhor que tenha isto como seguimento...e que siga duma
introdução extendida

At the core of the synthesis process, and of the developed synthetiser \synname,
is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
resource management. It connects the concepts seen in
background~\ref{sec:background} -- a type system
consisting of inference rules in sequent calculus for linear logic connectives;
with resource management, and focusing, as formalized techniques to,
respectively, solve resource related non-determinism, and to eliminate from the
proof search don't-care non-determinism.  Guided solely by this system's formal
representation\ref{sec:formal_system}, we can (and did) create a very robust
synthetiser for a select part of the linear types that's able to synthetise
programs successfully from e.g. the major linear logic equivalences. For example,
given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
the program $\lambda a .\ \lambda b .$ let $c \tensor d$ =
$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.

@@

Having said that, we won't allow the use of the recursive call in $\Gamma$ right
away. If we did, every synth goal would have a trivial proof (a non-terminating
function that just calls itself) that could even shadow other valid, relevant,
solutions.  Instead, our framework allows the use of recursion only after having
deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
invariant: the recursive hypothesis can only be \textbf{used in recursive branches of
ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
hypothesis as arguments. To illustrate, having $\textrm{data List} =
\textrm{Nil}\ \vert\ \textrm{Cons}\ (A * \textrm{List})$, recursion is only
"available" when deconstructing ($\textrm {List} \vdash C$), and only in
the \emph{Cons} branch. % (TODO: rewrite? O bold é meio estranho não?)

@@

\section{Architecture}

@@

\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
the same premises more than once. Since these proof attempts might be expensive
performance-wise, and we can rely on the program's determinism (equal calls to
synthetise result in equal outcomes), we add to our synthetiser memoization, an
optimization technique that stores function calls and returns cached values for
equal inputs. For our synthetiser, two calls are equivalent when contexts
$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
the goal, are equivalent. Because comparing all of these is slow, our
implementation hashes the relevant values and uses the result as the key.
Furthermore, a branch we want to memoize might fail, not providing any result --
we also want to record synthesis failure, as it happens more often than
successful results. We capture failure of a logict computation using the $ifte$
function and represent failure or success information with the Maybe type. We
ran benchmarks to measure the impact of memoization on the performance. It's
especially noticeable in synthesis with multiple functions in the unrestricted
context, or more complex specifications:
%
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Goal & Time + $\sigma$ & TimeM + $\sigma$ \\
        \hline
        insertWithKey & 13.31 s $\pm$ 593.4 ms & 2.667 s $\pm$ 284.3 ms \\
        \hline
    \end{tabular}
\end{center}
%
The details of the memoization are very subject to change

@@

\mypara{Evaluation} Validation and evaluation of our work is fairly
easy. Simple questions such as \emph{does it typecheck}? \emph{can
  we synthesize}? \emph{how fast}?  are the core of our evaluation. In
more concrete terms, we can compare our times to state of the art
synthetizers, present a list of working examples of programs synthesized
by our framework, and corresponding benchmarks.

On the eventual addition of recursive functions, refinement types, and
interactive synthesis, we'll consider, respectively, other examples to
showcase recursive functions synthesis, updated times for benchmarks
with refinement types, and a display of the possible user interaction
in the user process, and the resulting function following it.

@@

\section{Evaluation}

@@

>>>>

\usemintedstyle{manni}

% TODO: Temos de ter o título assim? ficava mais fixe sem o número e sem o
% supervisor :P
\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita - Nº 55902 \\
    Supervisor: Bernardo Toninho}
\date{July 2021}
\institute{NOVA School of Science and Technology}

@@

% General motivation?
Program synthesis is an automated or semi-automated process of deriving a
program, i.e.~generating code, from a high-level specification. Synthesis can be
seen as a means to improve a programmer's productivity and program correctness
(e.g. through suggestion and autocompletion).
%
Specifications can take many forms: natural language~\cite{chen2021evaluating},
examples~\cite{DBLP:conf/popl/FrankleOWZ16}, polymorphic refinement
types~\cite{DBLP:conf/pldi/PolikarpovaKS16}, graded linear
types~\cite{DBLP:conf/lopstr/HughesO20}, \emph{etc.}, but, regardless the kind of
specification, program synthetisers have to deal with two main inherent sources
of complexity -- searching over a vast space of (potentially) valid programs,
and interpreting user intent.

Synthesis is said to be \emph{type-driven} when it uses types as a form of program
specification and produces an expression whose type matches the specification.
Type-driven synthesis frameworks usually leverage rich types as a way to make
specifications more expressive and prune the valid programs search space, while
maintaining a ``familiar'' specification interface (types) for the user.
%
Richer type systems allow for more precise types, which can statically eliminate
various kinds of logical errors by making certain invalid program states
ill-typed (e.g.,~a ``null aware'' type system will ensure at compile-time that
you cannot dereference a null-pointer).
%
% However, they can also be a burden -- the whole point of these type systems is
% to ensure that \emph{fewer} programs are deemed well-typed, which can pose
% additional challenges to the development process.
%
For instance, the type $\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow
\mathsf{Int}$ specifies a (curried) function that takes two integers and
produces an integer. Viewed as a specification, it is extremely imprecise (there
are an infinite number of functions that satisfy this specification).  However,
the richer type $(x{:}\mathsf{Int}) \rightarrow (y{:}\mathsf{Int}) \rightarrow
\{z{:}\mathsf{Int} \mid z = x+y\}$ very precisely specifies a function that
takes integer arguments $x$ and $y$ and returns an integer $z$ that is the sum
of $x$ and $y$.

The focus of our work is on type-driven synthesis where specifications take the
form of linear types.
%
% TODO: Deixar isto para mais à frente?  More concretely, we will explore the
% proof-theoretic foundations of linear types via the Curry-Howard
% correspondence with linear logic propositions, viewing program synthesis as
% proof search in linear logic.  (Said correspondence maps \emph{types} to
% \emph{propositions in a logic}, and \emph{programs} to \emph{proofs}).
%
Linear types differ from more traditional types in that they constrain resource
usage in programs by \emph{statically} limiting the number of times certain
resources can be used during their lifetime (linear resources must be used
\emph{exactly once}).
%
They can be applied to resource-aware programming such as concurrent programming
(e.g. session types for message passing
concurrency~\cite{DBLP:journals/mscs/CairesPT16}), memory-management
(e.g.~Rust's ownership types), safely updating-in-place mutable
structures~\cite{Bernardy_2018}, enforcing protocols for external \textsc{api}s~\cite{Bernardy_2018}, to name a few.
% TODO: Repensar esta parte dos exemplos de aplicação e de como podemos
% relacionar o nosso trabalho com futuras aplicações nestes campos

Despite their long-known potential (\emph{linear types can change the
world!}\cite{Wadler90lineartypes}, \cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and,
primarily, strong proof-theoretic
foundations \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00}, synthesis of linear functional programs
from linear types hasn't been previously explored in an extended context (e.g.
with recursion, polymorphism, ...).
%TODO: Sinto que faltam algumas palavras aqui..

In this paper we present a framework for synthesis of linear types extended with
recursive algebraic data types, parametric polymorphism and refinements.
% TODO: extensible? proving feasability of... ? showing potential for ... ?

First, we discuss the use of linear types as specifications, the outline of the
synthesis process, and what that entails for the synthetiser -- ilustrated with
concrete results (Section~\ref{sec:overview}). Then, we present the
algorithm and formal system driving the synthesis
(Section~\ref{sec:formal_system}). Aftwerwards, we describe the architecture 
of our synthetiser named \synname, and examine technical details and challenges of the
implementation (Section~\ref{sec:architecture}). Finally, we evaluate our work
through expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks
(i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).

% TODO: Why linear types?

%%

%%
%This work explores the synthesis of linear functional
%programs from types based on linear logic propositions (i.e.~linear
%types) by leveraging the Curry-Howard correspondence.

@@

@@

\section{Overview}\label{sec:overview}

The \synname\ synthetiser combines the features of linear types, recursion,
parametric polymorphism, recursive polymorphic? algebraic data types, and
refinement types.

The \synname\ language is an extension of the linear $\lambda$-calculus.

... in proof search, our programs are correct by construction -- if
we were able to synth it, then it typechecks.

Linear types provide a more certain specification, when compared to simple
traditional types. By requiring some types to be used linearly (consumed) we can
assure, for example, the deconstruction of an ADT. Given the goal $\bang(a \lolli b)
\lolli List\ a \lolli List\ b$, the program $\lambda x. \lambda y. Nil$ wouldn't
be valid. To synthetise a valid program, $List\ a$
would have to be deconstructed, and its constructor arguments used, resulting
in the map function without relying on any additional information:

\begin{minted}{haskell}
map c d = let !e = c in
    case d of
          Nil  -> Nil
        | Cons f ->
            let g*h = f in Cons (e g, map (!e) h);
\end{minted}

At the core of the synthesis process, and of the developed synthetiser \synname,
is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
resource management -- a type system consisting of inference rules in sequent
calculus for linear logic connectives; with resource management, and focusing,
as formalized techniques to, respectively, solve resource related
non-determinism, and to eliminate from the proof search don't-care
non-determinism.  Guided solely by this system's formal
representation\ref{sec:formal_system}, we can (and did) create a very robust
synthetiser for a select part of the linear types that's able to synthetise
programs successfully from e.g. the major linear logic equivalences. For
example, given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B)
\lolli C$, the program $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$
$c$ $d$ (commonly known as \emph{uncurry}) is synthetised.

@@

Having said that, we won't allow the use of the recursive call in $\Gamma$ right
away. If we did, every synth goal would have a trivial proof (a non-terminating
function that just calls itself) that could even shadow other valid, relevant,
solutions.  Instead, our framework allows the use of recursion only after having
deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
invariant: the recursive hypothesis can only be \textbf{used in recursive branches of
ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
hypothesis as arguments. To illustrate, having \mintinline{haskell}{data List =
Nil | Cons (A * List)}, recursion is only
"available" when deconstructing ($\textrm {List} \vdash C$), and only in
the \emph{Cons} branch. % (TODO: rewrite? O bold é meio estranho não?)

@@

\section{Architecture}\label{sec:architecture}

@@

\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
the same premises more than once. Since these proof attempts might be expensive
performance-wise, and we can rely on the program's determinism (equal calls to
synthetise result in equal outcomes), we add to our synthetiser memoization, an
optimization technique that stores function calls and returns cached values for
equal inputs. For our synthetiser, two calls are equivalent when contexts
$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
the goal, are equivalent. Because comparing all of these is slow, our
implementation hashes the relevant values and uses the result as the key.
Furthermore, a branch we want to memoize might fail, not providing any result --
we also want to record synthesis failure, as it happens more often than
successful results. We capture failure of a logict computation using the $ifte$
function and represent failure or success information with the Maybe type. We
ran benchmarks to measure the impact of memoization on the performance. It's
especially noticeable in synthesis with multiple functions in the unrestricted
context, or more complex specifications:
%
\begin{center}
    \begin{tabular}{ |c|c|c|c|c| }
        \hline
        Goal & Time + $\sigma$ & TimeM + $\sigma$ \\
        \hline
        insertWithKey & 13.31 s $\pm$ 593.4 ms & 2.667 s $\pm$ 284.3 ms \\
        \hline
    \end{tabular}
\end{center}
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality for polymorphic types.

@@

% \mypara{Evaluation} Validation and evaluation of our work is fairly
% easy. Simple questions such as \emph{does it typecheck}? \emph{can
%   we synthesize}? \emph{how fast}?  are the core of our evaluation. In
% more concrete terms, we can compare our times to state of the art
% synthetizers, present a list of working examples of programs synthesized
% by our framework, and corresponding benchmarks.

% On the eventual addition of recursive functions, refinement types, and
% interactive synthesis, we'll consider, respectively, other examples to
% showcase recursive functions synthesis, updated times for benchmarks
% with refinement types, and a display of the possible user interaction
% in the user process, and the resulting function following it.

@@

\section{Evaluation}\label{sec:evaluation}

@@

Synthesis of
